Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-09 paper/main.tex
Updates

	Causal discovery has received significant attention in the Directed
	Acyclic Graphs (DAGs) literature, leading to the development of
	numerous automated algorithms for learning DAGs from data. However,
	their adoption in applied domains remain limited, as researchers often
	prefer to construct DAGs manually based on domain knowledge. This
	preference arises due to several practical challenges with automated
	algorithms, such as their tendency to make obvious errors and output
	Markov Equivalence Classes (MECs) rather than a DAG. To assist
	researchers in constructing DAGs manually, we propose an iterative DAG
	construction approach that combines domain knowledge with data-driven
	insights. Our method leverages Conditional Independence (CI) testing to
	iteratively identify variable pairs where an edge is either missing or
	superfluous. Based on this information, we can choose to add missing
	edges with appropriate orientation based on domain knowledge or remove
	unnecessary ones. We also give a method to rank potential new edges
	based on their impact on the overall model fit. Empirical results show
	that this iterative approach achieves performance comparable to
	automated algorithms if the orientation of new edges is correctly
	determined in at least one out of two cases. Additionally, we show
	that, in the absence of an expert, a Large Language Model (LLM) can be
	used to determine edge orientations. We provide an implementation of
	our approach as both a webtool: <redacted for review> and a Python
	package <redacted for review>.

@@

In this paper, we propose a method that leverages this CI testing approach to
assist researchers during the DAG construction process, rather than just
validating the model. Our approach follows an iterative model-building strategy
that integrates domain expertise with CI testing on data. At each iterations,
we use CI testing to identify pairs of variables where an edge is either
missing or superfluous. These pairs are then presented to the expert, who
determines whether to add or remove these edges. The orientation of the edge
when adding is determined based on the domain expertise. As at any iteration,
multiple changes to the DAG are possible. We also give a way to rank potential
new edges additions to the DAG using effect size measures for CI tests to
address the most severe violations of CI tests first. As a result this approach
acts like a guide to the expert when manually constructing their model while
ensuring that the model is consistent with the data.

@@

\subsection{Integrating domain knowledge in causal discovery}
Multiple ways to integrate, originally using forbidden and required edges. Or
temporal algorithms by specifying causal tiers.

\subsection{Measures of Conditional Association}
Our DAG structure learning approach relies on using CI tests and a measure of
conditional association between variables -- also known as partial association
-- between variables. Specifically, firstly we are interested in testing a CI
test: $ X \ci Y \bm{Z} $, i.e., whether the variables $ X $ and $ Y $ are
independent when conditioned on $ \bm{Z} $. Secondly, we are interested in
quantifying the strength of relationship between $ X $ and $ Y $ conditioned on
$ \bm{Z} $. We can quantify this using the effect size measures of the CI
tests. In this section, we give an overview of some of the commonly used CI
tests for various data types and their effect size measures.

@@

@@

% \begin{figure}
% 	\centering
% 	\begin{subfigure}{0.5\textwidth}
% 		\includegraphics[scale=0.25]{../../presentations/2024_05_das/2.png}
% 	\end{subfigure}%
% 	\begin{subfigure}{0.5\textwidth}
% 		\includegraphics[scale=0.25]{../../presentations/2024_05_das/5.png}
% 	\end{subfigure}
% 	\caption{Screenshots of the web-tool. \todo[inline]{Insert screenshots of the web-tool}}
% \end{figure}

>>>>

	Causal discovery has received significant attention in the Directed
	Acyclic Graphs (DAGs) literature, leading to the development of
	numerous automated algorithms for learning DAGs from data. However,
	their adoption in applied domains remain limited, as researchers often
	prefer to construct DAGs manually based on domain knowledge. This
	preference arises due to several practical challenges with automated
	algorithms, such as their tendency to make obvious errors and output
	Markov Equivalence Classes (MECs) rather than a DAG. To assist
	researchers in constructing DAGs manually, we propose an iterative 
	structure learning approach that combines domain knowledge with data-driven
	insights. Our method leverages Conditional Independence (CI) testing to
	iteratively identify variable pairs where an edge is either missing or
	superfluous. Based on this information, we can choose to add missing
	edges with appropriate orientation based on domain knowledge or remove
	unnecessary ones. We also give a method to rank potential new edges
	based on their impact on the overall model fit. Empirical results show
	that this iterative approach achieves performance comparable to
	automated algorithms if the orientation of new edges is correctly
	determined in at least one out of two cases. Additionally, we show
	that, in the absence of an expert, a Large Language Model (LLM) can be
	used to determine edge orientations. We provide an implementation of
	our approach as both a webtool: <redacted for review> and a Python
	package <redacted for review>.

@@

In this paper, we propose a structure learning method that leverages this CI
testing approach to assist researchers during the DAG construction process,
rather than just validating the model. Our approach follows an iterative
model-building strategy that integrates domain expertise with CI testing on
data. At each iterations, we use CI testing to identify pairs of variables
where an edge is either missing or superfluous. These pairs are then presented
to the expert, who determines whether to add or remove these edges. The
orientation of the edge when adding is determined based on the domain
expertise. As at any iteration, multiple changes to the DAG are possible. We
also give a way to rank potential new edges additions to the DAG using effect
size measures for CI tests to address the most severe violations of CI tests
first. As a result this approach acts like a guide to the expert when manually
constructing their model while ensuring that the model is consistent with the
data.

@@

\subsection{CI Tests and Measures of Conditional Association}
Our DAG structure learning approach relies on using CI tests and a measure of
conditional association between variables -- also known as partial association
-- between variables. Specifically, firstly we are interested in testing a CI
test: $ X \ci Y \bm{Z} $, i.e., whether the variables $ X $ and $ Y $ are
independent when conditioned on $ \bm{Z} $. Secondly, we are interested in
quantifying the strength of relationship between $ X $ and $ Y $ conditioned on
$ \bm{Z} $. We can quantify this using the effect size measures of the CI
tests. In this section, we give an overview of some of the commonly used CI
tests for various data types and their effect size measures.

@@

\subsection{Improving DAGs using CI Testing}

@@
